Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-09-03 Manuscript/manuscript.tex

The linear decrease of the headgroup order parameters, quantified in previous 
section, can be used to measure the amount of bound aqueous cations in lipid 
bilayers~\cite{seelig87,catte16}. The headgroup order parameter response 
to NaCl and CaCl$_2$ concentrations from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
The simulation results of Lipid14 with \AA{}qvist ions~\cite{aqvist90} (from Ref. \citenum{catte16})
are shown together with the results from Lipid14 simulation
with ions by Dang~\cite{smith94,chang1999,dang2006} and from ECC-lipids simulations with ECC-ions.

As already discussed in a recent work by NMRlipids project~\cite{catte16},
the headgroup order parameters are almost unaffected by addition of NaCl
in Lipid14 simulations with \AA{}qvist ion model due to a very weak binding of Na$^+$
in PC bilayer. This is in agremeent with experimental data, in contrast
to almost all the other lipid models, which significantly overestimate
Na$^+$ binding in PC bilayers~\cite{catte16}. Despite the good performance
with NaCl, the response of headgroup order parameters to CaCl$_2$ concentration are
significantly overestimated in Lipid14 simulations with \AA{}qvist ion model,
see Fig.~\ref{OrderParameterCHANGESnewMODELS} and Ref.~\citenum{catte16}.

Wide range of data presented in previous work \cite{catte16} suggested
that CaCl$_2$ binding behaviour cannot be corrected by only improving
ion models. This was not, however, tested specifically for Lipid14 model, so we
ran also Lipid14 simulations with ions by Dang~\cite{smith94,chang1999,dang2006} 
and ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014},
which have more realistic bulk behaviour than \AA{}qvist model~\cite{aqvist90}.
In line with previous work, the headgroup order parameters response
to CaCl$_2$ concentration was overestimated also with these ion models,
as seen in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.

Applying ECC correction also to lipids, introduced in section \ref{section:ecc},
brings the order parameter response to CaCl$_2$ concentration to a
good agreement with experiments, as shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
This is a significant improvement over previously introduced models,
which all overestimate the head group order parameter response to CaCl$_2$
concentration \cite{catte16}. The good agreement with experiments indicates
that the ECC corrected lipid model has a sufficient accuracy for realistic
studies of details of ions binding to lipid bilayers.

@@

Ion density profiles along membrane normal 
from different simulations are shown in Fig.~\ref{fig:cacl-dens}.
\todo{This paragraph is to be finalized when we have the surface excess analysis
  done by using Eq.~\ref{surfexcess}.}
This is directly related to the amount of surface excess charge 
as shown along with ion density profiles in  Fig.~\ref{fig:cacl-dens}.
The density profiles from simulations with original Lipid14 \cite{dickson14}
and Dang ions \cite{smith94,chang1999,dang2006} 
show a pronounced peak in the position of the phosphate moieties of POPC. 
The use of ECC-ion model~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}
or the ion model by \AA{qvist}~\cite{aqvist90}
along with Lipid14 does not significantly change it 
(Headgroup order parameter responses for these models in SI).
The new ECC-lipids model with ECC-ions~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}
exhibits on the other hand smaller density in this region 
suggesting overall weaker binding of cations (Fig.~\ref{OrderParameterCHANGESnewMODELS}). 
It is then likely that the models that overestimate head group order parameter response~\cite{catte16}
also overestimate the amount of bound cations.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    Relative surface excess for the respective simulations are: 
    Lipid14 with \AA{}qvist ions $\Gamma_i^w = XX nm^{-3}$; 
    Lipid14 with Dang ions $\Gamma_i^w = 0.3 nm^{-3}$; 
    ECC-lipids with ECC-ions $\Gamma_i^w = 0.07 nm^{-3}$. 
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-} and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration of the ions in water is 350~mM in all systems
    presented here. 
    }
  \todo{Calculate the surface excess for L14+Aqvist sim.} \\
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{Resuls with Lipid14+Dang to be added once done -- running with larger box.}
\end{figure}

>>>>

The linear decrease of the headgroup order parameters, quantified in previous 
section, can be used to measure the amount of bound aqueous cations in lipid 
bilayers~\cite{seelig87,catte16}. The headgroup order parameter response 
to NaCl and CaCl$_2$ concentrations from experiments (DPPC \cite{akutsu81} 
and POPC \cite{altenbach84}) and different simulation models for POPC
are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.
%The simulation results of Lipid14 with \AA{}qvist ions~\cite{aqvist90} (from Ref. \citenum{catte16})
%are shown together with the results from Lipid14 simulation
%with ions by Dang~\cite{smith94,chang1999,dang2006} and from ECC-lipids simulations with ECC-ions.

As already discussed previously~\cite{catte16}, NaCl has a negligible effect 
on the headgroup order parameters with submolar 
concentrations due to the very weak binding in PC bilayers. 
In contrast to almost all the other simulation models, this behaviour 
is reproduced by Lipid14 model with \AA{}qvist ions \cite{catte16}. 
The headgroup order parameter response to CaCl$_2$ concentration was,
however, overestimated with the same combination of force field
parameters~\cite{catte16}. The results were not improved when
ion models by Dang et al.~\cite{smith94,chang1999,dang2006} or
ECC-ions~\cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
with more realistic bulk behaviour were used, as seen in 
in Figs. \ref{OrderParameterCHANGESnewMODELS} and \ref{??} (in SI), respectively.
\todo{Add OP-response of Lipid14+ECC-ions plot in SI}.
Together with extensive data in previous work \cite{catte16}, 
the results suggest that CaCl$_2$ binding behaviour in PC bilayers
cannot be corrected by only improving the ion models.

Significant improvement is achieved when the developed ECC-lipid model 
is used. The headgroup order parameter changes as a function of CaCl$_2$
concentration model show a good agreement between ECC-lipid model and 
experiments in Fig. \ref{OrderParameterCHANGESnewMODELS}.
Ion binding affinity is quantified by using the relative surface
excess between ions and water from Eq.~\ref{surfexcess} and  
density profiles along membrane normal shown in Fig.~\ref{fig:cacl-dens}.
%This is directly related to the amount of surface excess charge 
%as shown along with ion density profiles in  Fig.~\ref{fig:cacl-dens}.
The relative surface excess for ECC-lipid model with ECC-ions $\Gamma_i^w = 0.07 nm^{-3}$
is significantly smaller than for Lipid14 with \AA{}qvist $\Gamma_i^w = XX nm^{-3}$
and Dang ions $\Gamma_i^w = 0.3 nm^{-3}$.
%The density profiles from simulations with original Lipid14 \cite{dickson14}
%and Dang ions \cite{smith94,chang1999,dang2006} 
%show a pronounced peak in the position of the phosphate moieties of POPC. 
%The use of ECC-ion model~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}
%or the ion model by \AA{qvist}~\cite{aqvist90}
%along with Lipid14 does not significantly change it 
%(Headgroup order parameter responses for these models in SI).
%The new ECC-lipids model with ECC-ions~\cite{Jungwirth2017,Jungwirth2015,kohagen14,kohagen16}
%exhibits on the other hand smaller density in this region 
%suggesting overall weaker binding of cations (Fig.~\ref{OrderParameterCHANGESnewMODELS}). 
%It is then likely that the models that overestimate head group order parameter response~\cite{catte16}
%also overestimate the amount of bound cations.
\todo{This paragraph is to be finalized when we have the surface excess analysis
  done by using Eq.~\ref{surfexcess}.}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=9.0cm,angle=0]{../Fig/CAdensities2.eps}
  \caption{\label{fig:cacl-dens}
    Number density of \ce{Ca^{2+}} and \ce{Cl^-} as a function of membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    Densities of~\ce{Cl^-} and water are divided with 2 and 200, respectively, to visalize them
    with the same scale as \ce{Ca^{2+}}. The molar concentration of the ions in water is 350~mM in all systems
    presented here. 
    }
  \todo{Calculate the surface excess for L14+Aqvist sim.} \\
  \todo{PAVEL: draw phosphate position with its variance, add water density (scaled) and include the number of$\Gamma$-surface access.} \\
  \todo{JOE: Change the figure so that it contains a membrane background} \\
  \todo{Resuls with Lipid14+Dang to be added once done -- running with larger box.} \\
  \todo{The current data for Dang simulation seems to contain more ions than others.}
\end{figure}

The good agreement with experiments indicates
that the ECC corrected lipid model has a sufficient accuracy for realistic
studies of details of ions binding to lipid bilayers.

@@
